\scuba: iterative map-making with the Sub-Millimetre User Reduction Facility

### 2012-09-05 smurf_paper.tex
move pre-processing section before model components
section 3 on map-maker workings mostly done

Significant experimentation with different models for noise sources
during \scuba\ commissioning required a highly flexible software
framework, and a configurable user interface. To achieve this goal,
while minimizing development time, it was decided to build SMURF as a
Starlink package \citep{2009ASPC..411..418J}, which provides access to
a large suite of libraries (including a commanding and messaging
interface, astrometric coordinate conversions and generation of
standard WCS information, file formats etc.). Furthermore, Starlink is
already used extensively at the Joint Astronomy Centre (host of the
JCMT) for many other systems, which helps with
interoperability. Though originally written primarily in FORTRAN, many
of the core Starlink libraries are now ported into native C, or else
have a C interface. It was therefore decided to develop SMURF in C
(rather than C++ which would have required adding further dependencies
to Starlink), although we have taken an object-oriented approach. For
example, all of the data for a given noise model is stored in a C
structure, and there is a standard interface for all functions that
handle models (member data and functions for the class,
respectively). In this way it is easy to extend SMURF with new
models. Parallelisation is incorporated in the most time-consuming
low-level routines using threads (e.g. performing Fast Fourier
Transforms, re-gridding the data), usually handling either independent
blocks of bolometers, or blocks of time, in each thread, depending on
the nature of the calculation. \textbf{... something about scalability
...}

@@

First, the data files are read from disk, concatenated into contiguous
arrays in memory, and run through a pre-processing stage that performs
several operations including spike and sudden level step removal to
enforce continuity in the time-series. Then, the iterative process
within the dashed box begins. It is generally assumed that the
dominant noise component is the common-mode, which is presumably
dominated by large variations in the fridge temperature and sky noise
(as argued in Section~\ref{sec:pca}). This signal is estimated as the
average signal observed by all of the bolometers at each time step,
and then removed from all of the bolometers. Once this large signal
has been removed, we apply the extinction correction, which is
determined completely from external measurements
\citep[see][]{dempsey2012}. Next, residual noise that differs from
bolometer-to-bolometer (primarily low-frequency) is removed, usually
using a high-pass filter. The resulting ``clean'' time-series are then
re-gridded to produce a map estimate using nearest-neighbour
sampling. Since each map pixel is calculated as the average of many
bolometer samples, it is fundamentally much less noisy than the raw
bolometer time-series. We can therefore project the map back into the
time domain (i.e. scanning the array across the current estimate of
the map), and create approximately noiseless estimates of what each
bolometer would have seen in the absence of noise sources, and remove
that from the time-series. This final step in the iteration gives a
residual signal (with low-frequency and astronomical noise components
removed) that may be used to measured the white noise level in each
detector, and also to estimate an approximate $\chi^2$ for the model
to track convergence. As we will discuss in the next sections, each
signal component estimate is biased slightly due to the presence of
signal from other components, so the entire process is
iterated. However, before the next iteration begins, we first undo the
multiplicative factors introduced by the extinction correction, and
also relative gains that may optionally be determined in the
common-mode estimation phase.

Map estimation is accomplished using a nearest-neighbour resampling of
the data onto a pre-defined map grid. For the $i$th map pixel
$\mathbf{m}(x_i,y_i)$, the brightness is estimated as the weighted
average of the bolometer data samples $\mathbf{b}_j$ that land within
that pixel (from any bolometer or point in time),
%
\begin{equation}
  \mathbf{m}(x_i,y_i) = \frac{\sum_j \mathbf{w}_j \mathbf{b}_j }
                             { \sum_j \mathbf{w}_j } .
\end{equation}
%
For the initial iteration the weights $\mathbf{w}_j$ are set to 1, but
subsequently they are set to $1/\sigma_j^2$, the estimated inverse
variance expected from the bolometer white noise levels, as discussed
in Section~\ref{sec:noi}. This weighting scheme is sensible provided
that the bolometer data have no correlated (i.e., low-frequency)
noise.

In addition to map estimation, the \model{AST} model can also be used
to perform map-based despiking of the time-series
(Section~\ref{sec:mapdespike}), and to apply constraints to the map to
improve convergence (Section~\ref{sec:converge}).

\subsection{Convergence tests and model degeneracies}
\label{sec:converge}

@@

In this section we describe a number of data preparation steps that
preceed the calculation of model components as described in the
previous section.

@@

The highest useful frequency in the nominally 200\,Hz-sampled
\scuba\ data is that which corresponds to the smallest angular scale
that the instrument is sensitive to. As mentioned earlier, the usual
rule-of-thumb for a Gaussian beam is to provide at least 3 samples for
each FWHM, or roughly 2.5\,arcsec for the 7.5\,arcsec
450\,\micron\ channel, and 5\,arcsec for the 14.5\,arcsec
850\,\micron\ channel. For a typical scan speed of
300\,arcsec\,s$^{-1}$, the maximum useful sample rate is therefore
about 120\,Hz at 450\,\micron, and 40\,Hz at 850\,\micron. In order to
save execution time and memory usage (both of which scale linearly
with data volume), it is clearly advantageous to re-sample the data to
these lower rates.

Similarly, from the point of view of retaining all of the useful
information, the size of the output map pixels need not be any smaller
than the FWHM/3 rule. Larger pixels may be selected, but at the
expense of retaining information at the diffraction limit of the
telescope.  \textbf{ Per: loss of information?}.

The default behaviour of SMURF is to select output map pixel sizes of
2 and 4\,arcsec at 450 and 850\,\micron, respectively (slightly
over-sampled). The slew speed of the telescope (approximately constant
for all standard scan modes) is then used to calculate the highest
useful frequency for the time-series given the map pixel size,
$f_\mathrm{max}$. The data are then re-sampled to this new rate prior
to map-making (as well as the data cleaning steps).

The method used by SMURF to down-sample simply averages together
multiple samples from the original time-series, $x_i$ to estimate the
lower-frequency output time-series, $y_j$. In general there will not
be an integer number of samples from $x_i$ in each of the $y_j$, so
fractional samples from $x_i$ are used at the $y_j$ time
boundaries. The algorithm is fast, since each sample in $x_i$ need
only be visited once. From a spectral point of view, this boxcar
average is equivalent to applying a sinc-function low-pass filter in
frequency space. Such behaviour is desirable, since it serves as an
anti-aliasing filter; most non-white features above the Nyquist
frequency in $y_j$ that could contaminate the output data are largely
removed.

As an alternative, we also investigated an algorithm in which the FFT
of $x_i$ is simply truncated to the target sample rate before
transforming back to the time domain (i.e., applying a hard-edged
low-pass filter). In practice the noise performance was
indistinguishable, and required slightly more execution time, using
this method compared to the boxcar method described above. It is
therefore not currently available to users.

@@

\item As a complete alternative to the iterative map-making procedure,
cleaning by principal component analysis (PCA) is available as an
experimental pre-processing option. As described in
Section~\ref{sec:pca}, a new set of basic vectors for the bolometer
data are found which diagonalize their covariance matrix. The
brightest components can then be subtracted from the data. A single
parameter specifies the threshold on the amplitude of the eigenvalues
to be removed, as a number of standard deviations away from the mean
value. Subarrays are cleaned independently once for the full length of
each continuousy chunk of data. Given the computational expense of
this method, and initial tests which showed little improvement over
simple high-pass filtering, this method has not yet been explored in
detail with \scuba\ data. However, since systematic effects seem to
come and go, it is possible that PCA could be useful for particular
data sets.

@@

Spikes of short duration and high amplitude are often seen in the time
series data. If not removed, they can cause ringing in the FLT model
(Section~\ref{sec:flt}). Two alternative approaches may be used to
remove these spikes. This section describes the detection and removal
of spikes within the time series of each bolometer, and
section~\ref{sec:mapdespike} describes the detection and removal of
spikes within the final map. In practice, map-based despiking
(Section~\ref{sec:mapdespike}) usually gives superior results, and so
time-domain despiking is usually switched off.

@@

Sudden steps can occur in the time stream data from each bolometer,
with the most likely cause being wraparound in the first stage squid
amplification, the details of which are given in \citet{holland2012}
\textbf{Per gave a technical explanation, but this is probably best in
another paper?}.

The black curves in Fig.~\ref{fig:steps1} show examples of such steps in
the time streams for two bolometers. If not removed, these steps can
cause severe ringing in the FLT model (Section~\ref{sec:flt}) and visible
streaks in the final map, corresponding to the paths of individual
bolometers over the sky.

@@

The evaluation of the \model{FLT} model (Section~\ref{sec:flt}) involves
taking the Fast Fourier Transform (FFT) of each bolometer time series.
Data that have been flagged as bad for any reason (for instance, due to
the presence of spikes, steps, or unusual common-mode signal) need to be
excluded from this FFT. For this reason, each contiguous block of bad
data samples is filled with artificial data before taking the FFT. A
least squares linear fit is performed to the 50 samples preceding the
block, and a similar fit is performed to the 50 samples following the
block. These are used to estimate the expected values at the start and
end of the block of bad values. The bad values are then replaced by
linear interpolation between the expected start and end values. Gaussian
noise is added with a standard deviation equal to the mean of the RMS
residuals in the two fits.

@@

\begin{itemize}
\item Apodisation: a number of samples at the start and end of each
bolometer time series are multiplied by a cosine function in order to
roll the data values off smoothly to zero. The default number of samples
modified at each end of the time series is given by half the ratio of the
sampling frequency to the lowest retained frequency. In addition, each
end of the time stream is padded with double this number of zeros. This
method is illustrated in Fig.~\ref{fig:pad2}.

@@

\item Padding with artificial data: instead of padding with zeros,
each time stream is padded with artificial data that connects the two
ends of the data stream smoothly and includes Gaussian noise. No
apodisation is performed. The number of samples of padding at each end
is again equal to the ratio of the sampling frequency to the lowest
retained frequency.  This is illustrated in Fig.~\ref{fig:pad1}. See
\citet{stompor2002} for a thorough discussion of this procedure within
the context of CMB map-making.

@@

\subsubsection{Flatfields}
\label{sec:flatfields}

TJ

Discuss heater ramps, normalizing off the common-mode signal
amplitude, astronomical sources, and the internal flatfield source.

\textit{Surely flatfields are discussed in the main \scuba\ paper so
don't really have any useful purpose here since they aren't
technically part of the map-maker.}

@@

TJ

Measuring the noise power spectra and throwing away outliers. Data
flagged bad by the DA system.

\subsubsection{Map-based despiking}
\label{sec:mapdespike}

It is possible to use the map estimate (Section~\ref{sec:ast}) to
provide additional despiking of the time-series. Unlike the
time-domain despiker (Section~\ref{sec:timedespike}), this calculation
uses the scatter in the population of samples used to estimate the map
pixel values (from different times and bolometers) to reject
outliers. This method is more robust against false-positive detections
of bright/compact astronomical sources, since transient features in the
time-series are unlikely to occur by chance whenever a bolometer
crosses a specific location on the sky, whereas real astronomical
sources are at fixed spatial locations.

The estimated variance, $\mathbf{v_p}(x_i,y_i)$, of the normalized
weighted samples that land in the $i$th map pixel is simply the biased
weighted sample variance (i.e., the variance map value multiplied by
the number of samples)
%
\begin{equation}
  \mathbf{v_p}(x_i,y_i) = N_i \mathbf{v}(x_i,y_i).
\end{equation}

In order to compare the weighted \emph{differences} between the
samples and the map values, [$\mathbf{b}_j - \mathbf{m}(x_i,y_i)$] to
$\mathbf{v_p}$, they must be scaled appropriately. We define a
normalized difference, $\mathbf{d}_j$, in such a way that the variance
of this new variable gives the weighted sample variance of the
underlying data points:
%
\begin{eqnarray}
  \frac{\sum_j \mathbf{d}_j^2}{N} &=&
  \frac{\sum_j \mathbf{w}_j [\mathbf{b}_j - \mathbf{m}(x_i,y_i)]^2}
       {\sum_k \mathbf{w}_k} \\
   \Rightarrow \mathbf{d}_j^2 &=& \frac{N \mathbf{w}_j [\mathbf{b}_j -
       \mathbf{m}(x_i,y_i)]^2}{ \sum_k \mathbf{w}_k} .
\end{eqnarray}
%
The map-based despiker flags those $\mathbf{d}_j$ that are further
than some threshold number of standard deviations,
$\sqrt{\mathbf{v_p}}$, away from zero so that they are not used in
subsequent iterations.

>>>>

Significant experimentation with different models for noise sources
during \scuba\ commissioning required a highly flexible software
framework, and a configurable user interface. To achieve this goal,
while minimizing development time, it was decided to build SMURF as a
Starlink package \citep{2009ASPC..411..418J}, which provides access to
a large suite of libraries (including a commanding and messaging
interface, astrometric coordinate conversions and generation of
standard WCS information, file formats etc.). Furthermore, Starlink is
already used extensively at the Joint Astronomy Centre (host of the
JCMT) for many other systems, which helps with
interoperability. Though originally written in FORTRAN, many of the
core Starlink libraries are now ported to native C, or at least have a
C interface. It was therefore decided to develop SMURF in C as well
(rather than C++, for example, which would have required adding
further dependencies to Starlink), although we have taken an
object-oriented approach. For example, all of the data for a given
noise model is encapsulated in a C structure, and there is a standard
interface for all functions that handle models (member data and
functions for the class, respectively). In this way it is easy to
extend SMURF with new models. Parallelisation is incorporated in the
most time-consuming low-level routines using threads (e.g. performing
Fast Fourier Transforms, re-gridding the data), usually handling
either independent blocks of bolometers, or blocks of time, in each
thread, depending on the nature of the
calculation. \textbf{... something about scalability ...}

@@

@@

Prior to executing the iterative part of the SMURF algorithm, the data
must undergo several pre-processing steps. First, the data files are
read into memory and concatenated into continuous arrays (\scuba\ data
are broken up and written to disk every $\sim$30\,s). As the data are
loaded, the flatfield correction is also applied \citep[see
Section~2.1 in][]{dempsey2012}. Next, a series of data cleaning and
filtering procedures are applied; these include the removal of large
glitches that may hinder the iterative solution from converging, or
simply tasks that do not need to be iterated.

@@

The highest useful frequency in the nominally 200\,Hz-sampled \scuba\
data is that which corresponds to the smallest angular scale that the
instrument is sensitive to. As mentioned earlier, the usual
rule-of-thumb for a Gaussian beam is to provide at least 3 samples for
each FWHM, or roughly 2.5\,arcsec for the 7.5\,arcsec 450\,\micron\
channel, and 5\,arcsec for the 14.5\,arcsec 850\,\micron\ channel. For
a typical scan speed of 300\,arcsec\,s$^{-1}$, the maximum useful
sample rate is therefore about 120\,Hz at 450\,\micron, and 40\,Hz at
850\,\micron. In order to save execution time and memory usage (both
of which scale linearly with data volume), it is clearly advantageous
to re-sample the data to these lower rates. In practice, the default
map pixel sizes are set to 2 and 4\,arcsec at 450 and 850\,\micron,
respectively (slightly over-sampled), and down-sampling occurs as the
data are loaded to match this spatial sampling given the slew speed.

The method used by SMURF to down-sample is to average together
multiple samples from the original time-series, $x_i$ to estimate the
lower-frequency output time-series, $y_j$. In general there will not
be an integer number of samples from $x_i$ in each of the $y_j$, so
fractional samples from $x_i$ are used at the $y_j$ time
boundaries. The algorithm is fast, since each sample in $x_i$ need
only be visited once. From a spectral point of view, this boxcar
average is equivalent to applying a sinc-function low-pass filter in
the frequency domain. Such behaviour is desirable since it serves as
an anti-aliasing filter; most non-white features above the Nyquist
frequency in $y_j$ that could contaminate the output data are largely
removed.

As an alternative, we also investigated an algorithm in which the FFT
of $x_i$ is simply truncated to the target sample rate before
transforming back to the time domain (i.e., applying a hard-edged
low-pass filter). In practice the noise performance was
indistinguishable, and required slightly more execution time using
this method. It is therefore not currently available to users.

@@

\item As an alternative to the iterative map-making procedure,
cleaning by principal component analysis (PCA) is available as an
experimental pre-processing option. As described in
Section~\ref{sec:pca}, a new set of basic vectors for the bolometer
data are found which diagonalize their covariance matrix. The
brightest components can then be subtracted from the data. A single
parameter specifies the threshold on the amplitude of the eigenvalues
to be removed, as a number of standard deviations away from the mean
value. Subarrays are cleaned independently once for the full length of
each continuousy chunk of data. Given the computational expense of
this method, and initial tests which showed little improvement over
simple high-pass filtering, this method has not yet been explored in
detail with \scuba\ data. However, since systematic effects seem to
come and go, it is possible that PCA could be useful for particular
data sets.

@@

Spikes of short duration and high amplitude are often seen in the time
series data. If not removed, they can cause ringing when filtering the
data. Two alternative approaches may be used to remove these
spikes. This section describes the detection and removal of spikes
within the time series of each bolometer, and Section~\ref{sec:ast}
describes the iterative detection and removal of spikes as part of map
estimation. In practice, map-based despiking usually gives superior
results, and so time-domain despiking is switched off by default.

@@

Sudden steps can occur in the time stream data from each bolometer,
with the most likely cause being cosmic ray events \citep[see
Section~3.5.3 in][]{holland2012}. The black curves in
Fig.~\ref{fig:steps1} show examples of such steps in the time streams
for two bolometers. If not removed, these steps can cause severe
ringing in the FLT model (Section~\ref{sec:flt}) and visible streaks
in the final map, corresponding to the paths of individual bolometers
over the sky.

@@

SMURF uses FFTs of the bolometer data extensively for filtering.  Data
that have been flagged as bad for any reason (for instance, due to the
presence of spikes, steps, or unusual common-mode signal) need to be
excluded from this FFT. For this reason, each contiguous block of bad
data samples is filled with artificial data before taking the FFT. A
least squares linear fit is performed to the 50 samples preceding the
block, and a similar fit is performed to the 50 samples following the
block. These are used to estimate the expected values at the start and
end of the block of bad values. The bad values are then replaced by
linear interpolation between the expected start and end
values. Gaussian noise is added with a standard deviation equal to the
mean of the RMS residuals in the two fits.

@@

\begin{itemize}
\item Apodisation: a number of samples at the start and end of each
bolometer time series are multiplied by a cosine function in order to
roll the data values off smoothly to zero. The default number of
samples modified at each end of the time series is given by half the
ratio of the sampling frequency to the lowest retained frequency. In
addition, each end of the time stream is padded with double this
number of zeros. This method is illustrated in
Fig.~\ref{fig:pad2}. This method is not used by default as it reduces
the amount of data available for the map, and can significantly hinder
very short observations (e.g. of calibrators when focussing the
telescope).

@@

\item Padding with artificial data: instead of padding with zeros,
each time stream is padded with artificial data that connects the two
ends of the data stream smoothly and includes Gaussian noise. No
apodisation is performed. The number of samples of padding at each end
is again equal to the ratio of the sampling frequency to the lowest
retained frequency.  This is illustrated in Fig.~\ref{fig:pad1}. See
\citet{stompor2002} for a thorough discussion of this procedure within
the context of CMB map-making. This method is used by default.

@@

@@

Despite all of the cleaning operations described in the previous
sections, the data from a given bolometer may be unusable do to it
being poorly biased, having an incorrect flatfield correction applied,
or having some other particularly pathological noise
contamination. Most of these bolometers can be flagged simply by
identifying outliers in the bolometer noise distribution. By default,
SMURF measures the PSD of all bolometers between 2 and 10\,Hz, after
all other pre-processing steps have been run (identical to the
measurement in Section~\ref{sec:noi}). Despite having significant
low-frequency noise (and possibly bright astronomical source)
contamination, this higher-frequency portion of the PSD is generally
quite clean. Furthermore, even if there is contamination, the purpose
of this measurement is to identify outliers, rather than provide a
meaningful absolute measurement of a given bolometer's white noise
level. Both high and low (e.g., due to an incorrect and very small
flatfield correction being applied) outliers from the centre of the
distribution are flagged. The measurement is based on the logarithm of
the noise values, and iterative clipping of the mean and standard
deviation are used, to reduce the impact of large outliers.

 and run through a pre-processing stage that performs
several operations including spike and sudden level step removal to
enforce continuity in the time-series. Then, the iterative process
within the dashed box begins. It is generally assumed that the
dominant noise component is the common-mode, which is presumably
dominated by large variations in the fridge temperature and sky noise
(as argued in Section~\ref{sec:pca}). This signal is estimated as the
average signal observed by all of the bolometers at each time step,
and then removed from all of the bolometers. Once this large signal
has been removed, we apply the extinction correction, which is
determined completely from external measurements
\citep[see][]{dempsey2012}. Next, residual noise that differs from
bolometer-to-bolometer (primarily low-frequency) is removed, usually
using a high-pass filter. The resulting ``clean'' time-series are then
re-gridded to produce a map estimate using nearest-neighbour
sampling. Since each map pixel is calculated as the average of many
bolometer samples, it is fundamentally much less noisy than the raw
bolometer time-series. We can therefore project the map back into the
time domain (i.e. scanning the array across the current estimate of
the map), and create approximately noiseless estimates of what each
bolometer would have seen in the absence of noise sources, and remove
that from the time-series. This final step in the iteration gives a
residual signal (with low-frequency and astronomical noise components
removed) that may be used to measured the white noise level in each
detector, and also to estimate an approximate $\chi^2$ for the model
to track convergence. As we will discuss in the next sections, each
signal component estimate is biased slightly due to the presence of
signal from other components, so the entire process is
iterated. However, before the next iteration begins, we first undo the
multiplicative factors introduced by the extinction correction, and
also relative gains that may optionally be determined in the
common-mode estimation phase.

Map estimation is accomplished using a nearest-neighbour resampling of
the data onto a pre-defined map grid with. For the $i$th map pixel
$\mathbf{m}(x_i,y_i)$, the brightness is estimated as the weighted
average of the bolometer data samples $\mathbf{b}_j$ that land within
that pixel (from any bolometer or point in time),
%
\begin{equation}
  \mathbf{m}(x_i,y_i) = \frac{\sum_j \mathbf{w}_j \mathbf{b}_j }
                             { \sum_j \mathbf{w}_j } .
\end{equation}
%
For the initial iteration the weights $\mathbf{w}_j$ are set to 1, but
subsequently they are set to $1/\sigma_j^2$, the estimated inverse
variance expected from the bolometer white noise levels, as discussed
in Section~\ref{sec:noi}. This weighting scheme is sensible provided
that the bolometer data have no correlated (i.e., low-frequency)
noise.

In addition to map estimation, the \model{AST} model can also be used
to perform map-based despiking of the time-series. Unlike the
time-domain despiker (Section~\ref{sec:timedespike}), this calculation
uses the scatter in the population of samples used to estimate the map
pixel values (from different times and bolometers) to reject
outliers. This method is more robust against false-positive detections
of bright/compact astronomical sources, since transient features in
the time-series are unlikely to occur by chance whenever a bolometer
crosses a specific location on the sky, whereas real astronomical
sources are at fixed spatial locations.

The estimated variance, $\mathbf{v_p}(x_i,y_i)$, of the normalized
weighted samples that land in the $i$th map pixel is simply the biased
weighted sample variance (i.e., the variance map value multiplied by
the number of samples)
%
\begin{equation}
  \mathbf{v_p}(x_i,y_i) = N_i \mathbf{v}(x_i,y_i).
\end{equation}
%
In order to compare the weighted \emph{differences} between the
samples and the map values, [$\mathbf{b}_j - \mathbf{m}(x_i,y_i)$] to
$\mathbf{v_p}$, they must be scaled appropriately. We define a
normalized difference, $\mathbf{d}_j$, in such a way that the variance
of this new variable gives the weighted sample variance of the
underlying data points:
%
\begin{eqnarray}
  \frac{\sum_j \mathbf{d}_j^2}{N} &=&
  \frac{\sum_j \mathbf{w}_j [\mathbf{b}_j - \mathbf{m}(x_i,y_i)]^2}
       {\sum_k \mathbf{w}_k} \\
   \Rightarrow \mathbf{d}_j^2 &=& \frac{N \mathbf{w}_j [\mathbf{b}_j -
       \mathbf{m}(x_i,y_i)]^2}{ \sum_k \mathbf{w}_k} .
\end{eqnarray}
%
The map-based despiker flags those $\mathbf{d}_j$ that are further
than some threshold number of standard deviations,
$\sqrt{\mathbf{v_p}}$, away from zero so that they are not used in
subsequent iterations.

A final option available as part of the \model{AST} model is to apply
constraints to the map to improve convergence. See the discussion in
Section~\ref{sec:converge}.

%-------------------------------------------------
\subsection{Convergence tests and model degeneracies}
\label{sec:converge}
%-------------------------------------------------
